---
title: "BIB Revision — Supplementary Analysis"
author: "Herdiantri Sufriyana, Yu-Wei Wu, Emily Chia-Yu Su"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
---

# Programming environment

```{r Load packages, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggpubr)
```

```{r Load functions, include=FALSE}
lapply(list.files("R/", full.names = TRUE), source)
```

```{r Load model registry, include=FALSE}
models <- read_csv("inst/extdata/model_registry.csv", show_col_types = FALSE)
```

```{r Session information, echo=FALSE}
sessionInfo()
```

```{r Helper function for computing AUPRC from ROC data, include=FALSE}
compute_auprc <- function(disc_rds, cal_rds) {
  roc <- disc_rds$plot$data
  d <- cal_rds$dist$data
  p <- d$n[d$obs == "1"] / sum(d$n)
  precision <- (roc$tpr * p) / (roc$tpr * p + (1 - roc$tnr) * (1 - p))
  recall <- roc$tpr
  precision[is.nan(precision)] <- 1
  ord <- order(recall, precision)
  recall <- recall[ord]
  precision <- precision[ord]
  idx <- !duplicated(recall)
  recall <- recall[idx]
  precision <- precision[idx]
  sum(diff(recall) * (precision[-1] + precision[-length(precision)]) / 2, na.rm = TRUE)
}

extract_metrics_for_model <- function(prefix, eval_dir, ont_name_file = NULL) {
  # Get all discrimination files (root + ontology)
  all_disc <- list.files(eval_dir, pattern = paste0("^", prefix, "_(root|ONT[0-9]+)_discrimination\\.rds$"), full.names = TRUE)

  results <- lapply(all_disc, function(f) {
    node <- str_extract(basename(f), "root|ONT[0-9]+")
    disc <- readRDS(f)
    cal_f <- sub("_discrimination\\.rds$", "_calibration.rds", f)
    cal <- if (file.exists(cal_f)) readRDS(cal_f) else NULL

    auroc <- disc$metrics$estimate
    auroc_ci <- disc$metrics$ci
    auprc <- if (!is.null(cal)) tryCatch(compute_auprc(disc, cal), error = function(e) NA) else NA
    rmse <- if (!is.null(cal)) cal$metrics$estimate[3] else NA
    rmse_ci <- if (!is.null(cal)) cal$metrics$ci[3] else NA

    # Map ONT ID to original
    original_id <- node
    if (!is.null(ont_name_file) && file.exists(ont_name_file) && node != "root") {
      ont_names <- read.csv(ont_name_file)
      ont_colon <- str_replace(node, "ONT", "ONT:")
      m <- ont_names[ont_names$new_oid == ont_colon, ]
      if (nrow(m) > 0) original_id <- m$old_oid[1]
    }

    data.frame(node = node, original_id = original_id,
               auroc = auroc, auroc_ci = auroc_ci,
               auprc = auprc, rmse = rmse, rmse_ci = rmse_ci)
  })

  do.call(rbind, results)
}
```

```{r Load all metrics for all models, include=FALSE}
disc_cal_data <- models |>
  rowwise() |>
  mutate(
    ont_name_file = paste0("data/ontology/", model_prefix, ".csv"),
    metrics_dev = list(extract_metrics_for_model(model_prefix, "data/eval_train/", ont_name_file)),
    metrics_test = list(extract_metrics_for_model(model_prefix, "data/eval/", ont_name_file))
  ) |>
  ungroup()
```

```{r Per-model tables, echo=FALSE, results='asis'}
fmt_auroc <- function(a, ci) sprintf("%.3f (%.3f, %.3f)", a, a - ci, a + ci)
fmt_val <- function(x) ifelse(is.na(x), "—", sprintf("%.3f", x))
fmt_rmse <- function(r, ci) ifelse(is.na(r), "—", sprintf("%.3f (±%.4f)", r, ci))

for (i in 1:nrow(disc_cal_data)) {
  m <- disc_cal_data[i, ]
  section_title <- paste0(m$pe_subtype, " — ", m$associated_factor,
                           " (Fig ", m$figure_ref, ", ", m$ontology_type, ")")
  table_caption <- paste0("Table ", i, ". AUROC, AUPRC, and calibration.")

  dev <- m$metrics_dev[[1]] |> arrange(node != "root", node)
  test <- m$metrics_test[[1]] |> arrange(node != "root", node)

  test_matched <- dev |>
    select(node) |>
    left_join(test |> select(node, auroc, auroc_ci, auprc, rmse, rmse_ci), by = "node", suffix = c("", "_test"))

  combined <- data.frame(Ontology = dev$original_id) |>
    cbind(
      data.frame(
        `AUC-ROC (Dev)` = fmt_auroc(dev$auroc, dev$auroc_ci),
        `AUC-ROC (Test)` = fmt_auroc(test_matched$auroc, test_matched$auroc_ci),
        `AUC-PRC (Dev)` = fmt_val(dev$auprc),
        `AUC-PRC (Test)` = fmt_val(test_matched$auprc),
        `RMSE (Dev)` = fmt_rmse(dev$rmse, dev$rmse_ci),
        `RMSE (Test)` = fmt_rmse(test_matched$rmse, test_matched$rmse_ci),
        check.names = FALSE
      )
    )

  cat(sprintf("\n# %s\n\n", section_title))

  print(
    combined |>
      kable(caption = table_caption, format = "html") |>
      kable_classic() |>
      scroll_box(height = "400px") |>
      column_spec(1:7, extra_css = "vertical-align:top;")
  )

  cat("\n\n")
}
```

# ROC curves

```{r ROC curves, echo=FALSE, fig.width=12, fig.height=8, warning=FALSE}
roc_plots <- models |>
  rowwise() |>
  mutate(
    disc = list(readRDS(paste0("data/eval/", model_prefix, "_root_discrimination.rds"))),
    auroc = disc$metrics$estimate,
    plot = list({
      df <- disc$plot$data
      ggplot(df, aes(x = 1 - tnr, y = tpr)) +
        geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
        geom_path(color = "steelblue", linewidth = 1) +
        labs(title = paste0("Fig ", figure_ref, ": ", model_id),
             subtitle = sprintf("AUC = %.3f", auroc),
             x = "1 - Specificity (FPR)", y = "Sensitivity (TPR)") +
        coord_equal() + theme_minimal() + theme(plot.title = element_text(size = 10))
    })
  ) |> pull(plot)
ggarrange(plotlist = roc_plots, ncol = 3, nrow = 3)
```

# Training-validation convergence

```{r Load convergence data, include=FALSE}
conv_data <- models |>
  rowwise() |>
  mutate(
    hist_file = paste0("data/history/", model_prefix, "_history.csv"),
    history = list(if (file.exists(hist_file)) read_csv(hist_file, show_col_types = FALSE) else NULL)
  ) |>
  ungroup()
```

```{r Convergence plots, echo=FALSE, fig.width=12, fig.height=8, warning=FALSE}
conv_plots <- conv_data |>
  filter(!map_lgl(history, is.null)) |>
  rowwise() |>
  mutate(plot = list({
    df <- history |> select(epoch, subset, total_loss)
    ggplot(df, aes(x = epoch, y = total_loss, color = subset)) +
      geom_line(linewidth = 0.8) +
      labs(title = paste0("Fig ", figure_ref, ": ", model_id),
           x = "Epoch", y = "Total loss", color = "Set") +
      scale_color_manual(values = c("training" = "steelblue", "validation" = "darkred")) +
      theme_minimal() + theme(plot.title = element_text(size = 10), legend.position = "bottom")
  })) |> pull(plot)
ggarrange(plotlist = conv_plots, ncol = 3, nrow = 3, common.legend = TRUE, legend = "bottom")
```

```{r Convergence AUC plots, echo=FALSE, fig.width=12, fig.height=8, warning=FALSE}
conv_auc_plots <- conv_data |>
  filter(!map_lgl(history, is.null)) |>
  rowwise() |>
  mutate(plot = list({
    df <- history |> select(epoch, subset, roc_auc)
    ggplot(df, aes(x = epoch, y = roc_auc, color = subset)) +
      geom_line(linewidth = 0.8) +
      labs(title = paste0("Fig ", figure_ref, ": ", model_id),
           x = "Epoch", y = "AUC-ROC", color = "Set") +
      scale_color_manual(values = c("training" = "steelblue", "validation" = "darkred")) +
      ylim(0, 1) +
      theme_minimal() + theme(plot.title = element_text(size = 10), legend.position = "bottom")
  })) |> pull(plot)
ggarrange(plotlist = conv_auc_plots, ncol = 3, nrow = 3, common.legend = TRUE, legend = "bottom")
```

# Threshold values and confusion matrices

Thresholds were selected from the development set by maximizing the net benefit
difference relative to the treat-all and treat-none references. Confusion matrix
metrics (PPV, TPR, NPV, TNR) are reported on the test set at the selected thresholds.

```{r Load thresholding and decision data, include=FALSE}
thr_dev <- models |>
  rowwise() |>
  mutate(thr = list(readRDS(paste0("data/eval_train/", model_prefix, "_root_thresholding.rds")))) |>
  ungroup()

thr_test <- models |>
  rowwise() |>
  mutate(thr = list(readRDS(paste0("data/eval/", model_prefix, "_root_thresholding.rds")))) |>
  ungroup()

dec_dev <- models |>
  rowwise() |>
  mutate(
    dec = list(readRDS(paste0("data/eval_train/", model_prefix, "_root_decision.rds"))),
    ta_intercept = dec$plot$layers[[3]]$data$intercept,
    ta_slope = dec$plot$layers[[3]]$data$slope
  ) |>
  ungroup()
```

```{r Combined threshold and confusion matrix table, echo=FALSE}
fmt_cm <- function(x, lb, ub) ifelse(is.nan(x) | is.na(x), "—", sprintf("%.1f%% (%.1f, %.1f)", x, lb, ub))

fmt_pair <- function(a_avg, a_lb, a_ub, b_avg, b_lb, b_ub) {
  paste0(fmt_cm(a_avg, a_lb, a_ub), " / ", fmt_cm(b_avg, b_lb, b_ub))
}

get_metric <- function(thr_rds, m) {
  list(avg = thr_rds$avg[thr_rds$metric == m],
       lb = thr_rds$lb[thr_rds$metric == m],
       ub = thr_rds$ub[thr_rds$metric == m])
}

rows <- list()
for (i in 1:nrow(models)) {
  m <- models[i, ]
  td <- thr_dev[i, ]
  tt <- thr_test[i, ]
  dd <- dec_dev[i, ]

  threshold <- td$thr[[1]]$ref_value[1]
  nb_model <- td$thr[[1]]$avg[td$thr[[1]]$metric == "nb"]
  nb_treat_all <- dd$ta_intercept + dd$ta_slope * threshold
  nb_ref <- max(nb_treat_all, 0)

  pairs <- list(
    list(label = "PPV/TPR", m1 = "ppv", m2 = "tpr"),
    list(label = "NPV/TNR", m1 = "npv", m2 = "tnr")
  )

  for (j in seq_along(pairs)) {
    p <- pairs[[j]]
    d1 <- get_metric(td$thr[[1]], p$m1); d2 <- get_metric(td$thr[[1]], p$m2)
    t1 <- get_metric(tt$thr[[1]], p$m1); t2 <- get_metric(tt$thr[[1]], p$m2)

    valid_dev <- is.finite(d1$lb) && d1$lb > 0 && is.finite(d2$lb) && d2$lb > 0
    valid_test <- is.finite(t1$lb) && t1$lb > 0 && is.finite(t2$lb) && t2$lb > 0
    valid <- if (valid_dev && valid_test) "Yes" else "No"

    rows[[length(rows) + 1]] <- data.frame(
      Model = if (j == 1) m$model_id else "",
      Figure = if (j == 1) m$figure_ref else "",
      Threshold = if (j == 1) sprintf("%.2f", threshold) else "",
      `NB Model` = if (j == 1) ifelse(is.nan(nb_model), "—", sprintf("%.4f", nb_model)) else "",
      `NB Ref` = if (j == 1) sprintf("%.4f", nb_ref) else "",
      Metrics = p$label,
      Dev = fmt_pair(d1$avg, d1$lb, d1$ub, d2$avg, d2$lb, d2$ub),
      Test = fmt_pair(t1$avg, t1$lb, t1$ub, t2$avg, t2$lb, t2$ub),
      Valid = valid,
      check.names = FALSE
    )
  }
}

do.call(rbind, rows) |>
  kable(caption = "Table 8. Threshold selection (dev set net benefit) and confusion matrix metrics (dev and test set, 95% CIs).", format = "html") |>
  kable_classic() |> column_spec(1:9, extra_css = "vertical-align:top;")
```

# Decision curves

Decision curves from the development set, where thresholds were selected.

```{r Decision curve plots, echo=FALSE, fig.width=12, fig.height=8, warning=FALSE}
dec_plots <- dec_dev |>
  rowwise() |>
  mutate(plot = list({
    df <- dec$plot$data
    ggplot(df, aes(x = th, y = nb)) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
      geom_abline(intercept = ta_intercept, slope = ta_slope,
                  linetype = "dotted", color = "darkred", linewidth = 0.8) +
      geom_line(color = "steelblue", linewidth = 1) +
      annotate("text", x = 0.02, y = ta_intercept + 0.005, label = "Treat all",
               color = "darkred", hjust = 0, size = 3) +
      annotate("text", x = 0.98, y = 0.005, label = "Treat none",
               color = "grey50", hjust = 1, size = 3) +
      labs(title = paste0("Fig ", figure_ref, ": ", model_id),
           x = "Threshold probability", y = "Net benefit") +
      xlim(0, 1) + theme_minimal() + theme(plot.title = element_text(size = 10))
  })) |> pull(plot)
ggarrange(plotlist = dec_plots, ncol = 3, nrow = 3)
```
