---
title: "BIB Revision — Supplementary Analysis"
author: "Herdiantri Sufriyana, Yu-Wei Wu, Emily Chia-Yu Su"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
---

# Programming environment

```{r Load packages, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggpubr)
```

```{r Load functions, include=FALSE}
lapply(list.files("R/", full.names = TRUE), source)
```

```{r Session information, echo=FALSE}
sessionInfo()
```

# Valid models

```{r Load model registry, include=FALSE}
models <- read_csv("inst/extdata/model_registry.csv", show_col_types = FALSE)
```

```{r Model registry table, echo=FALSE}
models |>
  select(-model_prefix, -ontology_type) |>
  kable(caption = "Table 1. The 7 validated pathogenesis models.", format = "html") |>
  kable_classic() |>
  column_spec(1:5, extra_css = "vertical-align:top;")
```

# Model discrimination (AUC-ROC)

```{r Load root discrimination, include=FALSE}
disc_data <- models |>
  rowwise() |>
  mutate(
    disc = list(readRDS(paste0("data/eval/", model_prefix, "_root_discrimination.rds"))),
    auroc = disc$metrics$estimate,
    auroc_ci = disc$metrics$ci
  ) |>
  ungroup()
```

## Root-level AUC-ROC

DI-VNN evaluates at specific ontology nodes, not root. Root-level AUC near 0.5
is expected — the model's value is in identifying *which* ontology nodes are
predictive.

```{r Root AUROC table, echo=FALSE}
disc_data |>
  transmute(
    `Model` = model_id,
    `PE Subtype` = pe_subtype,
    `Associated Factor` = associated_factor,
    `Figure` = figure_ref,
    `AUC-ROC` = sprintf("%.3f (%.3f, %.3f)", auroc, auroc - auroc_ci, auroc + auroc_ci)
  ) |>
  kable(caption = "Table 2. Root-level AUC-ROC.", format = "html") |>
  kable_classic() |>
  column_spec(1:5, extra_css = "vertical-align:top;")
```

## Ontology-level AUC-ROC summary

```{r Load ontology discrimination, include=FALSE}
ont_disc <- models |>
  rowwise() |>
  mutate(
    files = list(list.files("data/eval/", pattern = paste0("^", model_prefix, "_ONT[0-9]+_discrimination\\.rds$"), full.names = TRUE))
  ) |>
  unnest(files) |>
  rowwise() |>
  mutate(
    ontology = str_extract(files, "ONT[0-9]+"),
    d = list(readRDS(files)),
    auroc = d$metrics$estimate,
    auroc_ci = d$metrics$ci
  ) |>
  ungroup()
```

```{r Ontology AUROC summary table, echo=FALSE}
ont_disc |>
  group_by(model_id, pe_subtype, associated_factor, figure_ref) |>
  summarize(
    n = n(),
    min = min(auroc), max = max(auroc), median = median(auroc),
    .groups = "drop"
  ) |>
  transmute(
    Model = model_id, Figure = figure_ref,
    `N Ontologies` = n,
    `AUC-ROC Range` = sprintf("%.3f – %.3f", min, max),
    `Median` = sprintf("%.3f", median)
  ) |>
  kable(caption = "Table 3. Ontology-level AUC-ROC summary.", format = "html") |>
  kable_classic() |>
  column_spec(1:5, extra_css = "vertical-align:top;")
```

## Ontology-level detailed metrics

```{r Load ontology detail, include=FALSE}
ont_detail <- models |>
  rowwise() |>
  mutate(
    ont_name_file = paste0("data/ontology/", model_prefix, ".csv"),
    ont_names = list(if (file.exists(ont_name_file)) read.csv(ont_name_file) else NULL),
    files = list(list.files("data/eval/", pattern = paste0("^", model_prefix, "_ONT.+_discrimination\\.rds$"), full.names = TRUE))
  ) |>
  unnest(files) |>
  rowwise() |>
  mutate(
    ontology = str_extract(files, "ONT[0-9]+"),
    ont_id_colon = str_replace(ontology, "ONT", "ONT:"),
    original_id = {
      if (!is.null(ont_names)) {
        m <- ont_names[ont_names$new_oid == ont_id_colon, ]
        if (nrow(m) > 0) m$old_oid[1] else ontology
      } else ontology
    },
    d = list(readRDS(files)),
    auroc = d$metrics$estimate, auroc_ci = d$metrics$ci,
    thr_f = sub("_discrimination\\.rds$", "_thresholding.rds", files),
    thr = list(if (file.exists(thr_f)) readRDS(thr_f) else NULL),
    threshold = if (!is.null(thr)) thr$ref_value[1] else NA_real_,
    ppv = if (!is.null(thr)) thr$avg[thr$metric == "ppv"] else NA_real_,
    ppv_lb = if (!is.null(thr)) thr$lb[thr$metric == "ppv"] else NA_real_,
    ppv_ub = if (!is.null(thr)) thr$ub[thr$metric == "ppv"] else NA_real_,
    tpr = if (!is.null(thr)) thr$avg[thr$metric == "tpr"] else NA_real_,
    tpr_lb = if (!is.null(thr)) thr$lb[thr$metric == "tpr"] else NA_real_,
    tpr_ub = if (!is.null(thr)) thr$ub[thr$metric == "tpr"] else NA_real_,
    npv = if (!is.null(thr)) thr$avg[thr$metric == "npv"] else NA_real_,
    npv_lb = if (!is.null(thr)) thr$lb[thr$metric == "npv"] else NA_real_,
    npv_ub = if (!is.null(thr)) thr$ub[thr$metric == "npv"] else NA_real_,
    tnr = if (!is.null(thr)) thr$avg[thr$metric == "tnr"] else NA_real_,
    tnr_lb = if (!is.null(thr)) thr$lb[thr$metric == "tnr"] else NA_real_,
    tnr_ub = if (!is.null(thr)) thr$ub[thr$metric == "tnr"] else NA_real_,
    nb_model = if (!is.null(thr)) thr$avg[thr$metric == "nb"] else NA_real_,
    dec_f = sub("_discrimination\\.rds$", "_decision.rds", files),
    dec = list(if (file.exists(dec_f)) readRDS(dec_f) else NULL),
    nb_treat_all = {
      if (!is.null(dec) && length(dec$plot$layers) >= 3) {
        ta <- dec$plot$layers[[3]]$data
        ta$intercept + ta$slope * threshold
      } else NA_real_
    }
  ) |>
  ungroup()
```

```{r Ontology detail tables, echo=FALSE, results='asis'}
fmt <- function(x, lb, ub) ifelse(is.nan(x) | is.na(x), "—", sprintf("%.1f (%.1f, %.1f)", x, lb, ub))
fmt_nb <- function(x) ifelse(is.nan(x) | is.na(x), "—", sprintf("%.4f", x))

for (m in unique(ont_detail$model_id)) {
  sub <- ont_detail |> filter(model_id == m) |> arrange(desc(auroc))
  label <- paste0(sub$pe_subtype[1], " — ", sub$associated_factor[1], " (Fig ", sub$figure_ref[1], ")")
  cat(sprintf("\n### %s\n\n", label))
  tbl <- sub |> transmute(
    Ontology = original_id,
    `AUC-ROC` = sprintf("%.3f (±%.3f)", auroc, auroc_ci),
    Threshold = sprintf("%.2f", threshold),
    `NB Model` = fmt_nb(nb_model), `NB Treat-All` = fmt_nb(nb_treat_all),
    `PPV (%)` = fmt(ppv, ppv_lb, ppv_ub), `TPR (%)` = fmt(tpr, tpr_lb, tpr_ub),
    `NPV (%)` = fmt(npv, npv_lb, npv_ub), `TNR (%)` = fmt(tnr, tnr_lb, tnr_ub)
  )
  print(tbl |> kable(caption = label, format = "html") |> kable_classic() |>
    scroll_box(height = "400px") |> column_spec(1:9, extra_css = "vertical-align:top;"))
  cat("\n\n")
}
```

# ROC curves (root level)

```{r ROC curves, echo=FALSE, fig.width=12, fig.height=8, warning=FALSE}
roc_plots <- disc_data |>
  rowwise() |>
  mutate(plot = list({
    df <- disc$plot$data
    ggplot(df, aes(x = 1 - tnr, y = tpr)) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
      geom_path(color = "steelblue", linewidth = 1) +
      labs(title = paste0("Fig ", figure_ref, ": ", model_id),
           subtitle = sprintf("AUC = %.3f", auroc),
           x = "1 - Specificity (FPR)", y = "Sensitivity (TPR)") +
      coord_equal() + theme_minimal() + theme(plot.title = element_text(size = 10))
  })) |> pull(plot)
ggarrange(plotlist = roc_plots, ncol = 3, nrow = 3)
```

# Model calibration

```{r Load calibration, include=FALSE}
cal_data <- models |>
  rowwise() |>
  mutate(
    cal = list(readRDS(paste0("data/eval/", model_prefix, "_root_calibration.rds"))),
    intercept = cal$metrics$estimate[1],
    slope = cal$metrics$estimate[2],
    rmse = cal$metrics$estimate[3],
    rmse_ci = cal$metrics$ci[3]
  ) |>
  ungroup()
```

```{r Calibration table, echo=FALSE}
cal_data |>
  transmute(Model = model_id, Figure = figure_ref,
    Intercept = sprintf("%.3f", intercept),
    Slope = ifelse(is.na(slope), "—", sprintf("%.3f", slope)),
    RMSE = sprintf("%.3f (±%.4f)", rmse, rmse_ci)
  ) |>
  kable(caption = "Table 4. Root-level calibration metrics.", format = "html") |>
  kable_classic() |> column_spec(1:5, extra_css = "vertical-align:top;")
```

## Calibration plots

```{r Calibration plots, echo=FALSE, fig.width=12, fig.height=8, warning=FALSE}
cal_plots <- cal_data |>
  rowwise() |>
  mutate(plot = list({
    df <- cal$plot$data
    ggplot(df, aes(x = pred, y = true)) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
      geom_point(size = 3, color = "steelblue") +
      geom_errorbar(aes(ymin = lb, ymax = ub), width = 0.02, color = "steelblue") +
      labs(title = paste0("Fig ", figure_ref, ": ", model_id),
           subtitle = sprintf("RMSE = %.3f", rmse),
           x = "Predicted probability", y = "Observed frequency") +
      xlim(0, 1) + ylim(0, 1) + coord_equal() + theme_minimal() +
      theme(plot.title = element_text(size = 10))
  })) |> pull(plot)
ggarrange(plotlist = cal_plots, ncol = 3, nrow = 3)
```

## Prediction distributions

```{r Distribution plots, echo=FALSE, fig.width=12, fig.height=8, warning=FALSE}
dist_plots <- cal_data |>
  rowwise() |>
  mutate(plot = list({
    df <- cal$dist$data
    ggplot(df, aes(x = pred, y = n, fill = obs)) +
      geom_col(position = "dodge", width = 0.05) +
      labs(title = paste0("Fig ", figure_ref, ": ", model_id),
           x = "Predicted probability", y = "Count", fill = "Outcome") +
      xlim(0, 1) + theme_minimal() + theme(plot.title = element_text(size = 10))
  })) |> pull(plot)
ggarrange(plotlist = dist_plots, ncol = 3, nrow = 3)
```

# Decision curves

## Development set (threshold selection)

Thresholds were selected from the development set by maximizing the net benefit
difference relative to the treat-all and treat-none references.

```{r Load dev decision, include=FALSE}
dec_dev <- models |>
  rowwise() |>
  mutate(
    dec = list(readRDS(paste0("data/eval_train/", model_prefix, "_root_decision.rds"))),
    ta_intercept = dec$plot$layers[[3]]$data$intercept,
    ta_slope = dec$plot$layers[[3]]$data$slope
  ) |>
  ungroup()
```

```{r Dev decision curve plots, echo=FALSE, fig.width=12, fig.height=8, warning=FALSE}
dec_dev_plots <- dec_dev |>
  rowwise() |>
  mutate(plot = list({
    df <- dec$plot$data
    ggplot(df, aes(x = th, y = nb)) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
      geom_abline(intercept = ta_intercept, slope = ta_slope,
                  linetype = "dotted", color = "darkred", linewidth = 0.8) +
      geom_line(color = "steelblue", linewidth = 1) +
      annotate("text", x = 0.02, y = ta_intercept + 0.005, label = "Treat all",
               color = "darkred", hjust = 0, size = 3) +
      annotate("text", x = 0.98, y = 0.005, label = "Treat none",
               color = "grey50", hjust = 1, size = 3) +
      labs(title = paste0("Fig ", figure_ref, ": ", model_id, " (Dev set)"),
           x = "Threshold probability", y = "Net benefit") +
      xlim(0, 1) + theme_minimal() + theme(plot.title = element_text(size = 10))
  })) |> pull(plot)
ggarrange(plotlist = dec_dev_plots, ncol = 3, nrow = 3)
```

## Test set (validation)

```{r Load test decision, include=FALSE}
dec_test <- models |>
  rowwise() |>
  mutate(
    dec = list(readRDS(paste0("data/eval/", model_prefix, "_root_decision.rds"))),
    ta_intercept = dec$plot$layers[[3]]$data$intercept,
    ta_slope = dec$plot$layers[[3]]$data$slope
  ) |>
  ungroup()
```

```{r Test decision curve plots, echo=FALSE, fig.width=12, fig.height=8, warning=FALSE}
dec_test_plots <- dec_test |>
  rowwise() |>
  mutate(plot = list({
    df <- dec$plot$data
    ggplot(df, aes(x = th, y = nb)) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
      geom_abline(intercept = ta_intercept, slope = ta_slope,
                  linetype = "dotted", color = "darkred", linewidth = 0.8) +
      geom_line(color = "steelblue", linewidth = 1) +
      annotate("text", x = 0.02, y = ta_intercept + 0.005, label = "Treat all",
               color = "darkred", hjust = 0, size = 3) +
      annotate("text", x = 0.98, y = 0.005, label = "Treat none",
               color = "grey50", hjust = 1, size = 3) +
      labs(title = paste0("Fig ", figure_ref, ": ", model_id, " (Test set)"),
           x = "Threshold probability", y = "Net benefit") +
      xlim(0, 1) + theme_minimal() + theme(plot.title = element_text(size = 10))
  })) |> pull(plot)
ggarrange(plotlist = dec_test_plots, ncol = 3, nrow = 3)
```

# Threshold analysis

## Development set (threshold selection)

```{r Load dev thresholding, include=FALSE}
thr_dev <- models |>
  rowwise() |>
  mutate(thr = list(readRDS(paste0("data/eval_train/", model_prefix, "_root_thresholding.rds")))) |>
  ungroup()
```

```{r Dev thresholds table, echo=FALSE}
thr_dev_joined <- thr_dev |>
  left_join(dec_dev |> select(model_id, ta_intercept, ta_slope), by = "model_id")

thr_dev_joined |>
  rowwise() |>
  mutate(
    threshold = thr$ref_value[1],
    nb_model = thr$avg[thr$metric == "nb"],
    nb_treat_all = ta_intercept + ta_slope * threshold,
    nb_ref = max(nb_treat_all, 0),
    nb_diff = nb_model - nb_ref,
    ppv = thr$avg[thr$metric == "ppv"], ppv_lb = thr$lb[thr$metric == "ppv"], ppv_ub = thr$ub[thr$metric == "ppv"],
    tpr = thr$avg[thr$metric == "tpr"], tpr_lb = thr$lb[thr$metric == "tpr"], tpr_ub = thr$ub[thr$metric == "tpr"],
    npv = thr$avg[thr$metric == "npv"], npv_lb = thr$lb[thr$metric == "npv"], npv_ub = thr$ub[thr$metric == "npv"],
    tnr = thr$avg[thr$metric == "tnr"], tnr_lb = thr$lb[thr$metric == "tnr"], tnr_ub = thr$ub[thr$metric == "tnr"]
  ) |>
  ungroup() |>
  transmute(
    Model = model_id, Figure = figure_ref,
    Threshold = sprintf("%.2f", threshold),
    `NB Model` = ifelse(is.nan(nb_model), "—", sprintf("%.4f", nb_model)),
    `NB Ref` = sprintf("%.4f", nb_ref),
    `NB Diff` = ifelse(is.nan(nb_diff), "—", sprintf("%.4f", nb_diff)),
    PPV = ifelse(is.nan(ppv), "—", sprintf("%.1f%% (%.1f, %.1f)", ppv, ppv_lb, ppv_ub)),
    TPR = ifelse(is.nan(tpr), "—", sprintf("%.1f%% (%.1f, %.1f)", tpr, tpr_lb, tpr_ub)),
    NPV = ifelse(is.nan(npv), "—", sprintf("%.1f%% (%.1f, %.1f)", npv, npv_lb, npv_ub)),
    TNR = ifelse(is.nan(tnr), "—", sprintf("%.1f%% (%.1f, %.1f)", tnr, tnr_lb, tnr_ub))
  ) |>
  kable(caption = "Table 5. Development set: thresholds, net benefits (model vs. best reference), and performance metrics (95% CIs).", format = "html") |>
  kable_classic() |> column_spec(1:10, extra_css = "vertical-align:top;")
```

## Test set (validation)

```{r Load test thresholding, include=FALSE}
thr_test <- models |>
  rowwise() |>
  mutate(thr = list(readRDS(paste0("data/eval/", model_prefix, "_root_thresholding.rds")))) |>
  ungroup()
```

```{r Test thresholds table, echo=FALSE}
thr_test_joined <- thr_test |>
  left_join(dec_test |> select(model_id, ta_intercept, ta_slope), by = "model_id")

thr_test_joined |>
  rowwise() |>
  mutate(
    threshold = thr$ref_value[1],
    nb_model = thr$avg[thr$metric == "nb"],
    nb_treat_all = ta_intercept + ta_slope * threshold,
    nb_ref = max(nb_treat_all, 0),
    nb_diff = nb_model - nb_ref,
    ppv = thr$avg[thr$metric == "ppv"], ppv_lb = thr$lb[thr$metric == "ppv"], ppv_ub = thr$ub[thr$metric == "ppv"],
    tpr = thr$avg[thr$metric == "tpr"], tpr_lb = thr$lb[thr$metric == "tpr"], tpr_ub = thr$ub[thr$metric == "tpr"],
    npv = thr$avg[thr$metric == "npv"], npv_lb = thr$lb[thr$metric == "npv"], npv_ub = thr$ub[thr$metric == "npv"],
    tnr = thr$avg[thr$metric == "tnr"], tnr_lb = thr$lb[thr$metric == "tnr"], tnr_ub = thr$ub[thr$metric == "tnr"]
  ) |>
  ungroup() |>
  transmute(
    Model = model_id, Figure = figure_ref,
    Threshold = sprintf("%.2f", threshold),
    `NB Model` = ifelse(is.nan(nb_model), "—", sprintf("%.4f", nb_model)),
    `NB Ref` = sprintf("%.4f", nb_ref),
    `NB Diff` = ifelse(is.nan(nb_diff), "—", sprintf("%.4f", nb_diff)),
    PPV = ifelse(is.nan(ppv), "—", sprintf("%.1f%% (%.1f, %.1f)", ppv, ppv_lb, ppv_ub)),
    TPR = ifelse(is.nan(tpr), "—", sprintf("%.1f%% (%.1f, %.1f)", tpr, tpr_lb, tpr_ub)),
    NPV = ifelse(is.nan(npv), "—", sprintf("%.1f%% (%.1f, %.1f)", npv, npv_lb, npv_ub)),
    TNR = ifelse(is.nan(tnr), "—", sprintf("%.1f%% (%.1f, %.1f)", tnr, tnr_lb, tnr_ub))
  ) |>
  kable(caption = "Table 6. Test set: thresholds, net benefits (model vs. best reference), and performance metrics (95% CIs).", format = "html") |>
  kable_classic() |> column_spec(1:10, extra_css = "vertical-align:top;")
```
