---
title: "BIB Revision — Supplementary Analysis"
author: "Herdiantri Sufriyana, Yu-Wei Wu, Emily Chia-Yu Su"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
---

# Programming environment

```{r Load packages, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggpubr)
```

```{r Load functions, include=FALSE}
lapply(list.files("R/", full.names = TRUE), source)
```

```{r Load model registry, include=FALSE}
models <- read_csv("inst/extdata/model_registry.csv", show_col_types = FALSE)
```

```{r Session information, echo=FALSE}
sessionInfo()
```

# 1. AUROC

```{r Load discrimination, include=FALSE}
disc_data <- models |>
  rowwise() |>
  mutate(
    disc = list(readRDS(paste0("data/eval/", model_prefix, "_root_discrimination.rds"))),
    auroc = disc$metrics$estimate,
    auroc_ci = disc$metrics$ci
  ) |>
  ungroup()
```

```{r AUROC table, echo=FALSE}
disc_data |>
  transmute(
    Model = model_id, Figure = figure_ref, Subtype = pe_subtype,
    `Associated Factor` = associated_factor,
    `AUC-ROC` = sprintf("%.3f (%.3f, %.3f)", auroc, auroc - auroc_ci, auroc + auroc_ci)
  ) |>
  kable(caption = "Table 1. Root-level AUC-ROC (test set).", format = "html") |>
  kable_classic() |> column_spec(1:5, extra_css = "vertical-align:top;")
```

# 2. AUPRC

```{r Compute AUPRC, include=FALSE}
auprc_data <- models |>
  rowwise() |>
  mutate(
    disc = list(readRDS(paste0("data/eval/", model_prefix, "_root_discrimination.rds"))),
    cal = list(readRDS(paste0("data/eval/", model_prefix, "_root_calibration.rds"))),
    prevalence = {
      d <- cal$dist$data
      d$n[d$obs == "1"] / sum(d$n)
    },
    auprc = {
      roc <- disc$plot$data
      p <- prevalence
      precision <- (roc$tpr * p) / (roc$tpr * p + (1 - roc$tnr) * (1 - p))
      recall <- roc$tpr
      precision[is.nan(precision)] <- 1
      ord <- order(recall, precision)
      recall <- recall[ord]
      precision <- precision[ord]
      idx <- !duplicated(recall)
      recall <- recall[idx]
      precision <- precision[idx]
      sum(diff(recall) * (precision[-1] + precision[-length(precision)]) / 2, na.rm = TRUE)
    }
  ) |>
  ungroup()
```

```{r AUPRC table, echo=FALSE}
auprc_data |>
  transmute(
    Model = model_id, Figure = figure_ref,
    Prevalence = sprintf("%.3f", prevalence),
    `AUC-PRC` = sprintf("%.3f", auprc)
  ) |>
  kable(caption = "Table 2. Root-level AUC-PRC (test set). Low prevalence results in low baseline AUPRC.", format = "html") |>
  kable_classic() |> column_spec(1:4, extra_css = "vertical-align:top;")
```

# 3. Calibration

```{r Load calibration, include=FALSE}
cal_data <- models |>
  rowwise() |>
  mutate(
    cal = list(readRDS(paste0("data/eval/", model_prefix, "_root_calibration.rds"))),
    intercept = cal$metrics$estimate[1],
    slope = cal$metrics$estimate[2],
    rmse = cal$metrics$estimate[3],
    rmse_ci = cal$metrics$ci[3]
  ) |>
  ungroup()
```

```{r Calibration table, echo=FALSE}
cal_data |>
  transmute(
    Model = model_id, Figure = figure_ref,
    Intercept = sprintf("%.3f", intercept),
    Slope = ifelse(is.na(slope), "—", sprintf("%.3f", slope)),
    RMSE = sprintf("%.3f (±%.4f)", rmse, rmse_ci)
  ) |>
  kable(caption = "Table 3. Root-level calibration metrics (test set).", format = "html") |>
  kable_classic() |> column_spec(1:5, extra_css = "vertical-align:top;")
```

## Calibration plots

```{r Calibration plots, echo=FALSE, fig.width=12, fig.height=8, warning=FALSE}
cal_plots <- cal_data |>
  rowwise() |>
  mutate(plot = list({
    df <- cal$plot$data
    ggplot(df, aes(x = pred, y = true)) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
      geom_point(size = 3, color = "steelblue") +
      geom_errorbar(aes(ymin = lb, ymax = ub), width = 0.02, color = "steelblue") +
      labs(title = paste0("Fig ", figure_ref, ": ", model_id),
           subtitle = sprintf("RMSE = %.3f", rmse),
           x = "Predicted probability", y = "Observed frequency") +
      xlim(0, 1) + ylim(0, 1) + coord_equal() + theme_minimal() +
      theme(plot.title = element_text(size = 10))
  })) |> pull(plot)
ggarrange(plotlist = cal_plots, ncol = 3, nrow = 3)
```

## Prediction distributions

```{r Distribution plots, echo=FALSE, fig.width=12, fig.height=8, warning=FALSE}
dist_plots <- cal_data |>
  rowwise() |>
  mutate(plot = list({
    df <- cal$dist$data
    ggplot(df, aes(x = pred, y = n, fill = obs)) +
      geom_col(position = "dodge", width = 0.05) +
      labs(title = paste0("Fig ", figure_ref, ": ", model_id),
           x = "Predicted probability", y = "Count", fill = "Outcome") +
      xlim(0, 1) + theme_minimal() + theme(plot.title = element_text(size = 10))
  })) |> pull(plot)
ggarrange(plotlist = dist_plots, ncol = 3, nrow = 3)
```

# 4. ROC curves

```{r ROC curves, echo=FALSE, fig.width=12, fig.height=8, warning=FALSE}
roc_plots <- disc_data |>
  rowwise() |>
  mutate(plot = list({
    df <- disc$plot$data
    ggplot(df, aes(x = 1 - tnr, y = tpr)) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
      geom_path(color = "steelblue", linewidth = 1) +
      labs(title = paste0("Fig ", figure_ref, ": ", model_id),
           subtitle = sprintf("AUC = %.3f", auroc),
           x = "1 - Specificity (FPR)", y = "Sensitivity (TPR)") +
      coord_equal() + theme_minimal() + theme(plot.title = element_text(size = 10))
  })) |> pull(plot)
ggarrange(plotlist = roc_plots, ncol = 3, nrow = 3)
```

# 5. Confusion matrices

```{r Load test thresholding, include=FALSE}
thr_test <- models |>
  rowwise() |>
  mutate(thr = list(readRDS(paste0("data/eval/", model_prefix, "_root_thresholding.rds")))) |>
  ungroup()
```

```{r Confusion matrix table, echo=FALSE}
thr_test |>
  rowwise() |>
  mutate(
    threshold = thr$ref_value[1],
    ppv = thr$avg[thr$metric == "ppv"], ppv_lb = thr$lb[thr$metric == "ppv"], ppv_ub = thr$ub[thr$metric == "ppv"],
    tpr = thr$avg[thr$metric == "tpr"], tpr_lb = thr$lb[thr$metric == "tpr"], tpr_ub = thr$ub[thr$metric == "tpr"],
    npv = thr$avg[thr$metric == "npv"], npv_lb = thr$lb[thr$metric == "npv"], npv_ub = thr$ub[thr$metric == "npv"],
    tnr = thr$avg[thr$metric == "tnr"], tnr_lb = thr$lb[thr$metric == "tnr"], tnr_ub = thr$ub[thr$metric == "tnr"]
  ) |>
  ungroup() |>
  transmute(
    Model = model_id, Figure = figure_ref,
    Criteria = validation_criteria,
    Threshold = sprintf("%.2f", threshold),
    PPV = ifelse(is.nan(ppv), "—", sprintf("%.1f%% (%.1f, %.1f)", ppv, ppv_lb, ppv_ub)),
    TPR = ifelse(is.nan(tpr), "—", sprintf("%.1f%% (%.1f, %.1f)", tpr, tpr_lb, tpr_ub)),
    NPV = ifelse(is.nan(npv), "—", sprintf("%.1f%% (%.1f, %.1f)", npv, npv_lb, npv_ub)),
    TNR = ifelse(is.nan(tnr), "—", sprintf("%.1f%% (%.1f, %.1f)", tnr, tnr_lb, tnr_ub))
  ) |>
  kable(caption = "Table 4. Confusion matrix metrics at selected thresholds (test set, 95% CIs).", format = "html") |>
  kable_classic() |> column_spec(1:8, extra_css = "vertical-align:top;")
```

# 6. Training-validation convergence

```{r Load convergence data, include=FALSE}
conv_data <- models |>
  rowwise() |>
  mutate(
    hist_file = paste0("data/history/", model_prefix, "_history.csv"),
    history = list(if (file.exists(hist_file)) read_csv(hist_file, show_col_types = FALSE) else NULL)
  ) |>
  ungroup()
```

```{r Convergence plots, echo=FALSE, fig.width=12, fig.height=8, warning=FALSE}
conv_plots <- conv_data |>
  filter(!map_lgl(history, is.null)) |>
  rowwise() |>
  mutate(plot = list({
    df <- history |> select(epoch, subset, total_loss)
    ggplot(df, aes(x = epoch, y = total_loss, color = subset)) +
      geom_line(linewidth = 0.8) +
      labs(title = paste0("Fig ", figure_ref, ": ", model_id),
           x = "Epoch", y = "Total loss", color = "Set") +
      scale_color_manual(values = c("training" = "steelblue", "validation" = "darkred")) +
      theme_minimal() + theme(plot.title = element_text(size = 10), legend.position = "bottom")
  })) |> pull(plot)
ggarrange(plotlist = conv_plots, ncol = 3, nrow = 3, common.legend = TRUE, legend = "bottom")
```

```{r Convergence AUC plots, echo=FALSE, fig.width=12, fig.height=8, warning=FALSE}
conv_auc_plots <- conv_data |>
  filter(!map_lgl(history, is.null)) |>
  rowwise() |>
  mutate(plot = list({
    df <- history |> select(epoch, subset, roc_auc)
    ggplot(df, aes(x = epoch, y = roc_auc, color = subset)) +
      geom_line(linewidth = 0.8) +
      labs(title = paste0("Fig ", figure_ref, ": ", model_id),
           x = "Epoch", y = "AUC-ROC", color = "Set") +
      scale_color_manual(values = c("training" = "steelblue", "validation" = "darkred")) +
      ylim(0, 1) +
      theme_minimal() + theme(plot.title = element_text(size = 10), legend.position = "bottom")
  })) |> pull(plot)
ggarrange(plotlist = conv_auc_plots, ncol = 3, nrow = 3, common.legend = TRUE, legend = "bottom")
```

# 7. Threshold values

```{r Load dev thresholding, include=FALSE}
thr_dev <- models |>
  rowwise() |>
  mutate(thr = list(readRDS(paste0("data/eval_train/", model_prefix, "_root_thresholding.rds")))) |>
  ungroup()

dec_dev <- models |>
  rowwise() |>
  mutate(
    dec = list(readRDS(paste0("data/eval_train/", model_prefix, "_root_decision.rds"))),
    ta_intercept = dec$plot$layers[[3]]$data$intercept,
    ta_slope = dec$plot$layers[[3]]$data$slope
  ) |>
  ungroup()
```

Thresholds were selected from the development set by maximizing the net benefit
difference relative to the treat-all and treat-none references.

```{r Threshold values table, echo=FALSE}
thr_dev |>
  left_join(dec_dev |> select(model_id, ta_intercept, ta_slope), by = "model_id") |>
  rowwise() |>
  mutate(
    threshold = thr$ref_value[1],
    nb_model = thr$avg[thr$metric == "nb"],
    nb_treat_all = ta_intercept + ta_slope * threshold,
    nb_ref = max(nb_treat_all, 0),
    nb_diff = nb_model - nb_ref
  ) |>
  ungroup() |>
  transmute(
    Model = model_id, Figure = figure_ref,
    Threshold = sprintf("%.2f", threshold),
    `NB Model` = ifelse(is.nan(nb_model), "—", sprintf("%.4f", nb_model)),
    `NB Ref (max of treat-all, 0)` = sprintf("%.4f", nb_ref),
    `NB Diff` = ifelse(is.nan(nb_diff), "—", sprintf("%.4f", nb_diff))
  ) |>
  kable(caption = "Table 5. Threshold selection from development set: net benefit vs. best reference.", format = "html") |>
  kable_classic() |> column_spec(1:6, extra_css = "vertical-align:top;")
```

# 8. Decision curves

Decision curves from the development set, where thresholds were selected.

```{r Decision curve plots, echo=FALSE, fig.width=12, fig.height=8, warning=FALSE}
dec_plots <- dec_dev |>
  rowwise() |>
  mutate(plot = list({
    df <- dec$plot$data
    ggplot(df, aes(x = th, y = nb)) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
      geom_abline(intercept = ta_intercept, slope = ta_slope,
                  linetype = "dotted", color = "darkred", linewidth = 0.8) +
      geom_line(color = "steelblue", linewidth = 1) +
      annotate("text", x = 0.02, y = ta_intercept + 0.005, label = "Treat all",
               color = "darkred", hjust = 0, size = 3) +
      annotate("text", x = 0.98, y = 0.005, label = "Treat none",
               color = "grey50", hjust = 1, size = 3) +
      labs(title = paste0("Fig ", figure_ref, ": ", model_id),
           x = "Threshold probability", y = "Net benefit") +
      xlim(0, 1) + theme_minimal() + theme(plot.title = element_text(size = 10))
  })) |> pull(plot)
ggarrange(plotlist = dec_plots, ncol = 3, nrow = 3)
```
